Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-12 paper2.qmd
section:6; process:write

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from these processes.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the two hypothetical texts. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from the general form, each progressively incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and to "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference. Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. It begins by formulating a conceptual model represented by an SCM and a DAG. This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms. The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model. In particular, it enables the analysis of CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when researchers seek to conduct statistical inference or hypothesis testing (refer to @sec-theory-issue2). These methodological innovations potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical conclusions.

@@

<!-- 2. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two additional advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs--one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.

@@

For instance,

<!-- accounts for judge biases, captures the hierarchical structure of stimuli, incorporates measurement error in trait estimation and hypothesis testing, and accommodates violations of the equal dispersion assumption.  -->

<!-- 3. First research avenue -->
<!-- One promising research avenue involves investigating how the hierarchical structure of stimuli affects CJ experiment designs and hypotheses testing. One such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy. Due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation]. This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous. However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates. -->

<!-- One of these avenues could be about the implications of the hierarchical structure of the stimuli, not only for the design of CJ experiments, but also for the test of hypothesis.  -->
<!-- The avenues of research span the following:  -->
<!-- - assessing if the number of texts a researcher collects is sufficient. Due to efficiency of scoring, most of research only collects only one text. This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability. However, this cannot be known ad-hoc. The frameworks open the avenue to test this assumption -->
<!-- - Assessing hypothesis regarding the characteristics of the texts. Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->

<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
<!-- - does the judges have biases, and if so, do ad-hoc sta -->

<!-- (Is this a validity contention?) -->
<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp. 683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement. As @Kelly_et_al_2022 noted [pp. 683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->

@@

@@

@@

<!-- Challenges -->
<!-- 1. Epistemiology  -->
Despite its potential, this study presents several challenges that researchers must navigate. First, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate. In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest? However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces. Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.

<!-- 2. Research context is important  -->
Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts. We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions. That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.

<!-- 3. Causal inference knowledge -->
Third, applying this approach requires a solid understanding of Causal inference. For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity. Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.

<!-- 4. PPL and Bayesian Inference -->
Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods. This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand. While such translation is not exclusive to Bayesian methods--frequentist approaches can also be used--we argue that Bayesian methods offer a more accessible pathway,  as they reduce the need for additional expertise in optimization theory and techniques.

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement model, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical inferences derived from these processes.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the two hypothetical texts. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over A.

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference. Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

<!-- 4. General summary of what we have done. -->
Thus, to address the limitations of Thurstone's Case V and the BTL model--particularly their strong assumptions and the disconnect between trait measurement and hypothesis testing--this study extends Thurstone's general form using a systematic, integrated approach that combines Causal and Bayesian inference methods. The approach begins by formulating a conceptual model represented by an SCM and a DAG. This model integrates Thurstone's core theoretical principles--such as the discriminal processes (i.e., traits) of stimuli--alongside key CJ experimental design features, including judges' bias, sampling procedures, and comparison mechanisms.

The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model. This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2). In particular, the model accounts for judge biases, captures the hierarchical structure of stimuli, accommodates heterogeneity in discriminal dispersions, and incorporates measurement error in the hypothesis testing process. These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.

@@

<!-- 5. Are there any additional advantages? -->
Beyond these potential benefits, the approach offers two additional advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the analytic paradigm from passively accepting the assumptions of the BTL model to actively testing their fit with observed data. Together, these advantages create a principled framework for evaluating best practices in CJ experimental designs, one that better aligns with the demands of contemporary CJ assessment contexts [@Kelly_et_al_2022] and opens new avenues for research.

@@

<!-- 6. Which are these new avenues for research? -->
Five research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect the robustness of statistical inferences. Second, further work is needed to understand how judges' biases impact the number of comparisons required to achieve reliable trait estimates. Third, researchers should re-evaluate existing methods for identifying and excluding 'misfitting' judges and stimuli to ensure their effectiveness. Fourth, studies should explore how the hierarchical structure of stimuli influences both experimental design and hypothesis testing in CJ contexts. Finally,  studies should explore principled approaches for selecting contextually relevant variables.

<!-- 6.2 About the implications of the sampling and comparison mechanisms -->

[About the implications of the sampling and comparison mechanisms]{style="color:blue;"}

@@

<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp. 683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement. As @Kelly_et_al_2022 noted [pp. 683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->

<!-- 6.3 About judges biases, judgment discrepancies, and the number of judges and judgments -->

[About judges biases, judgment discrepancies, and the number of judges and judgments]{style="color:blue;"}

@@

<!-- presence of judges biases in the comparative judgment procedure. From this it derives research on:  -->
<!-- - does the judges have biases, and if so, do ad-hoc methods can identify them -->

<!-- 6.4 About removing misfit judges and stimuli  -->

[About removing misfit judges and stimuli]{style="color:blue;"}

<!-- 6.1 About the hierarchical structure of stimuli, number and predictors -->

[About the hierarchical structure of stimuli]{style="color:blue;"}

<!-- About the hierarchical structure of stimuli and its impact on CJ experiment design and hypothesis testing, one such study could examine whether collecting a single text per individual provides sufficient information to produce reliable trait estimates at different levels of the hierarchy. It has been highlighted that due to concerns about the feasibility of the comparison task [@Boonen_et_al_2020], many CJ studies assess traits using only one text per individual [@citation]. Although the literature have examples of collecting more than one stimuli per individual, their statistical handling still separates the trait measurement from the statistical inference process. This experimental design implicitly assumes that within-individual variability is small relative to between-individual differences--that is, texts produced by the same individual are relatively homogeneous. However, researchers rarely test this assumption, even though it directly influences decisions about the number of texts and participants needed for the experiment, ultimately affecting the reliability of the trait estimates. -->

<!-- The avenues of research span the following:  -->
<!-- - assessing if the number of texts a researcher collects is sufficient. Due to efficiency of scoring, most of research only collects only one text. This assumes that there is significant homogeneity between text of the same individual, compared to the between individual variability. However, this cannot be known ad-hoc. The frameworks open the avenue to test this assumption -->
<!-- - Assessing hypothesis regarding the characteristics of the texts. Before the literature did not have a direct and integrated way to assess these types of hypothesis, now you just need to collect data on the particular stimuli, and test if these variables reflect the expected effect -->

<!-- 6.5 About using Causal inference to carefully the most relevant variables for specific research contexts -->

[About revisit the approaches for selecting contextually relevant variables]{style="color:blue;"}

@@

<!-- 7. Challenges -->
<!-- 7.1 Epistemiology  -->
Despite its potential, this study presents several challenges that researchers must navigate. First, our approach does not establish a clear epistemological connection between the theoretical latent trait and its quantitative estimate. In other words, the approach does not provide a definitive answer to the fundamental question: How do we know that the estimated trait reflects the theoretical trait of interest? However, this limitation is not unique to our approach; it reflects a fundamental challenge all research involving latent constructs faces. Nevertheless, we argue that our approach offers methodological tools that enable researchers to design CJ experiments under conditions that support reliable and valid trait measurement.

<!-- 7.2 Research context is important  -->
Second, although initial simulation tests of the statistical model have yielded promising results (not shown), these findings may not generalize across research contexts. We have argued that the model's ability to accurately estimate traits and support valid statistical inferences depends heavily on the specific research context--particularly on whether the data and inferential goals align with the model's underlying assumptions. That said, our approach provides tools for researchers to test and evaluate these context-dependent CJ models directly.

<!-- 7.3 Causal inference knowledge -->
Third, applying this approach requires a solid understanding of Causal inference. For instance, throughout the study, we assume that predictor variables must be "relevant" to the research context, interpreting relevance as the inclusion of the predictor variables in a "sufficient adjustment set." However, we do not fully explore what this entails or its implications for model specification and validity. Nonetheless, we provide key references throughout the study to help motivated researchers explore these foundational concepts in greater depth.

<!-- 7.4 PPL and Bayesian Inference -->
Fourth and final, researchers must also be proficient in Probabilistic Programming Languages (PPLs) and Bayesian inference methods. This expertise is essential for translating context-specific conceptual models into bespoke statistical models that meaningfully address the research questions at hand. While such translation is not exclusive to Bayesian methods--frequentist approaches can also be used--we argue that Bayesian methods offer a more accessible pathway,  as they reduce the need for additional expertise in optimization theory and techniques.
